---
title: "Dynamic Structural Equation Model: Single-Factor, Two Periods"
author: "Greg Veramendi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Structural Equation Model: Single-Factor, Two Periods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

When the same latent construct is observed at two time points and you want
to estimate how the period-2 value depends on the period-1 value, you need
to combine two things:

1. **A measurement system** that is *invariant* across time (same loadings,
   thresholds / residual SDs across the two periods) so the factor scale is
   comparable, and
2. **A structural equation** linking the period-2 factor to the period-1
   factor: `f₂ = α + β·f₁ + ε`.

**factorana** supports this with a two-stage workflow:

- **Stage 1** fits a measurement model with the shared measurement slopes
  (loadings, sigmas) tied across periods but the measurement *intercepts*
  free to differ. The wave-1 intercepts pin the measurement scale at the
  level implied by the factor-identification convention `E[f₁] = 0`.
- **Stage 2** fixes the measurement system from Stage 1 and fits
  `factor_structure = "SE_linear"` to estimate the structural parameters
  `α`, `β`, and the structural residual variance `σ²_ε`.

This vignette walks through a simulated example where `α`, `β`, and
`σ²_ε` are known, shows the recovery, and explains one subtlety of the
workflow (why the wave-1 intercepts are the right ones to carry forward
into Stage 2).

## Simulate data

One latent construct, three linear indicators per period, two periods.
The DGP:

$$
f_1 \sim N(0,\sigma^2_{f_1}), \qquad
f_2 = \alpha + \beta\,f_1 + \varepsilon, \quad \varepsilon \sim N(0,\sigma^2_\varepsilon)
$$

$$
Y_{t,m} = \tau_m + \lambda_m\,f_t + u_{t,m}, \quad u_{t,m} \sim N(0,\sigma_m^2)
$$

with the measurement parameters `τ_m`, `λ_m`, `σ_m` shared across the
two periods.

```{r}
library(factorana)

set.seed(41)
n <- 1500

# Structural parameters
true_var_f1   <- 1.0
true_alpha    <- 0.4
true_beta     <- 0.6
true_sigma_e2 <- 0.5

# Shared measurement parameters
item_int   <- c(1.5, 1.0, 0.8)
item_load  <- c(1.0, 0.9, 1.1)   # first loading fixed to 1 for identification
item_sigma <- c(0.7, 0.75, 0.65)

f1  <- rnorm(n, 0, sqrt(true_var_f1))
eps <- rnorm(n, 0, sqrt(true_sigma_e2))
f2  <- true_alpha + true_beta * f1 + eps

gen_Y <- function(f, i) {
  item_int[i] + item_load[i] * f + rnorm(length(f), 0, item_sigma[i])
}

dat <- data.frame(
  intercept = 1,
  eval      = 1L,
  Y_t1_m1 = gen_Y(f1, 1), Y_t1_m2 = gen_Y(f1, 2), Y_t1_m3 = gen_Y(f1, 3),
  Y_t2_m1 = gen_Y(f2, 1), Y_t2_m2 = gen_Y(f2, 2), Y_t2_m3 = gen_Y(f2, 3)
)
head(dat[, -(1:2)], 3)
```

## Stage 1: measurement system with period-specific intercepts

We specify a 2-factor independent model. Factor 1 carries the period-1
indicators; factor 2 carries the period-2 indicators. We **tie the
loadings and sigmas** across periods via `equality_constraints`
(measurement invariance in slopes and noise), but **leave the intercepts
free and period-specific**.

```{r}
fm_stage1 <- define_factor_model(
  n_factors        = 2,
  n_types          = 1,
  factor_structure = "independent"
)

mk <- function(name, norm) {
  define_model_component(
    name = name, data = dat, outcome = name, factor = fm_stage1,
    covariates = "intercept", model_type = "linear",
    loading_normalization = norm, evaluation_indicator = "eval"
  )
}

comps_stage1 <- list(
  mk("Y_t1_m1", c(1, 0)),           # loading fixed to 1 on factor 1
  mk("Y_t1_m2", c(NA_real_, 0)),
  mk("Y_t1_m3", c(NA_real_, 0)),
  mk("Y_t2_m1", c(0, 1)),           # loading fixed to 1 on factor 2
  mk("Y_t2_m2", c(0, NA_real_)),
  mk("Y_t2_m3", c(0, NA_real_))
)

eq_constraints <- list(
  c("Y_t1_m2_loading_1", "Y_t2_m2_loading_2"),
  c("Y_t1_m3_loading_1", "Y_t2_m3_loading_2"),
  c("Y_t1_m1_sigma",     "Y_t2_m1_sigma"),
  c("Y_t1_m2_sigma",     "Y_t2_m2_sigma"),
  c("Y_t1_m3_sigma",     "Y_t2_m3_sigma")
)

ms_stage1 <- define_model_system(
  components           = comps_stage1,
  factor               = fm_stage1,
  equality_constraints = eq_constraints
)

ctrl <- define_estimation_control(n_quad_points = 8, num_cores = 1)

result_stage1 <- estimate_model_rcpp(
  model_system = ms_stage1,
  data         = dat,
  control      = ctrl,
  optimizer    = "nlminb",
  parallel     = FALSE,
  verbose      = FALSE
)
result_stage1$convergence   # 0 = strict convergence
```

Inspect the intercepts: the wave-1 intercepts should match the DGP
`τ_m` closely (because `E[f₁] = 0` by convention), while the wave-2
intercepts will be shifted by `λ_m · E[f₂]`.

```{r}
est <- result_stage1$estimates
cat(sprintf("%-6s  %8s  %8s  %8s\n", "m", "DGP tau", "wave-1", "wave-2"))
for (i in 1:3) {
  cat(sprintf("%-6d  %8.3f  %8.3f  %8.3f\n",
              i, item_int[i],
              est[paste0("Y_t1_m", i, "_intercept")],
              est[paste0("Y_t2_m", i, "_intercept")]))
}
```

Notice how the wave-2 intercepts are systematically higher than the
DGP — that's the mean shift `λ_m·E[f₂]` being absorbed. We will
**discard** those wave-2 intercepts and carry the **wave-1** intercepts
into Stage 2 in their place. That's what lets Stage 2 identify `α`
from the period-to-period mean shift in the observed `Y`.

## Why not pool the measurements naively?

A natural alternative would be to stack period-1 and period-2 rows into
one long-format dataset and fit a single-factor CFA with `2n` rows.
The resulting measurement intercepts would be the *pooled* means of
`Y` across both periods — biased by `λ_m · E[f₂]/2`, because the
pooled mean averages the period-1 and period-2 factor means. Plugging
those pooled intercepts into Stage 2 makes `α` under-estimate by
roughly `E[f₂]/2`. The 2-factor-with-tied-slopes setup above avoids
that bias while still pooling loadings and sigmas for precision.

## Build Stage 2's `previous_stage` using the wave-1 intercepts

We now construct a "dummy" 2-factor result that plugs the wave-1
intercepts into both factor slots. The construction is mechanical:

```{r}
get <- function(nm) unname(result_stage1$estimates[nm])

# Shared (tied) loadings and sigmas
lambda <- c(m2 = get("Y_t1_m2_loading_1"), m3 = get("Y_t1_m3_loading_1"))
sigma  <- c(m1 = get("Y_t1_m1_sigma"),
            m2 = get("Y_t1_m2_sigma"),
            m3 = get("Y_t1_m3_sigma"))
# Wave-1 intercepts — these will be used for BOTH factor slots
tau_w1 <- c(m1 = get("Y_t1_m1_intercept"),
            m2 = get("Y_t1_m2_intercept"),
            m3 = get("Y_t1_m3_intercept"))

# Rebuild a 2-factor independent model_system (identical structure to
# Stage 1) — but the parameter values we'll hand it use wave-1
# intercepts for BOTH t1 and t2 slots.
ms_dummy <- define_model_system(components = comps_stage1, factor = fm_stage1)

vals   <- c(factor_var_1 = get("factor_var_1"),
            factor_var_2 = get("factor_var_2"))
for (i in 1:3) {
  prefix <- paste0("Y_t1_m", i)
  vals[paste0(prefix, "_intercept")] <- tau_w1[[paste0("m", i)]]
  if (i > 1) vals[paste0(prefix, "_loading_1")] <- lambda[[paste0("m", i)]]
  vals[paste0(prefix, "_sigma")] <- sigma[[paste0("m", i)]]
}
for (i in 1:3) {
  prefix <- paste0("Y_t2_m", i)
  vals[paste0(prefix, "_intercept")] <- tau_w1[[paste0("m", i)]]   # wave-1
  if (i > 1) vals[paste0(prefix, "_loading_2")] <- lambda[[paste0("m", i)]]
  vals[paste0(prefix, "_sigma")] <- sigma[[paste0("m", i)]]
}

stage1_dummy <- list(
  model_system = ms_dummy,
  estimates    = vals,
  std_errors   = setNames(rep(0, length(vals)), names(vals)),
  convergence  = 0L,
  loglik       = 0.0
)
```

## Stage 2: `SE_linear`

With the dummy `previous_stage` built, Stage 2 is a single call:

```{r}
fm_stage2 <- define_factor_model(
  n_factors        = 2,
  n_types          = 1,
  factor_structure = "SE_linear"
)

ms_stage2 <- define_model_system(
  components     = list(),       # measurement components come from previous_stage
  factor         = fm_stage2,
  previous_stage = stage1_dummy
)

init_s2 <- initialize_parameters(ms_stage2, dat, verbose = FALSE)
init_s2$init_params["factor_var_1"]    <- get("factor_var_1")
init_s2$init_params["se_intercept"]    <- 0.0
init_s2$init_params["se_linear_1"]     <- 0.5
init_s2$init_params["se_residual_var"] <- 0.5

result_stage2 <- estimate_model_rcpp(
  model_system = ms_stage2,
  data         = dat,
  init_params  = init_s2$init_params,
  control      = ctrl,
  optimizer    = "nlminb",
  parallel     = FALSE,
  verbose      = FALSE
)
result_stage2$convergence
```

## Recovery

```{r}
est <- result_stage2$estimates
se  <- result_stage2$std_errors

tab <- data.frame(
  param = c("factor_var_1", "se_intercept", "se_linear_1", "se_residual_var"),
  true  = c(true_var_f1,     true_alpha,     true_beta,     true_sigma_e2),
  est   = unname(est[c("factor_var_1", "se_intercept", "se_linear_1", "se_residual_var")]),
  se    = unname(se[c("factor_var_1", "se_intercept", "se_linear_1", "se_residual_var")])
)
tab$z <- round((tab$est - tab$true) / tab$se, 2)
tab$est <- round(tab$est, 4)
tab$se  <- round(tab$se, 4)
knitr::kable(tab, digits = 4)
```

`se_intercept` (the structural α) recovers essentially exactly — that
is the parameter that motivated the whole workflow. `se_linear_1`
(the slope β), `se_residual_var` (σ²_ε), and `factor_var_1` also
recover within their standard errors.

## Where to go next

- `?define_factor_model` — supports `factor_structure = "SE_quadratic"`
  (adds a quadratic term `γ·f_1²` to the structural equation), useful for
  dynamics with a trap or threshold behaviour.
- `?define_model_system` — the `equality_constraints` argument is the
  key to the Stage 1 setup shown here; it accepts any number of tied
  parameter groups, each one a character vector of parameter names.
- `vignette("measurement_system", package = "factorana")` — basics of
  measurement-system estimation (no structural equation).
- `vignette("roy_model", package = "factorana")` — a different
  structural setting (discrete sector choice + continuous outcomes
  with a shared latent ability).
